A fast algorithm for global code motion of congruent instructions

### 2017-11-13 cgo/paper.tex
reformat

  for each CHI-nodes in the CFG:
    for each CHI-arg in CHI-node:
      Remove CHI-arg if it fails legality checks

  for each CHI-nodes in the CFG:
    if CHI  does not have missing entries:
      Perform code hoisting
      update data structures and statistics

@@

  for each VN in VNs
    for each instruction with the same VN
      Find its dominance-frontier (DF)
      Insert the PHI-node at DF

  for each PHI-nodes in the CFG:
    for each PHI-arg in PHI-node:
      Remove PHI-arg if it fails legality checks

  for each PHI-nodes in the CFG:
    if PHI does not have missing entries:
      Perform code sinking
      if a $\phi$ has same entries, reassign
      update data structures and statistics

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
For evaluation of \gcm{}, we built \SPEC{} with and without the \GCM{}
optimization. All the experiments were conducted on an \xlinux{} machine and at
-Ofast optimization level with inlining and loop unrolling disabled; inliner and
loop-unroller vary wildly with \GCM{} optimization so it was difficult to
visualize the impact. Each benchmark was run three times and the best result was
taken. We collected execution time and code-size which is listed in
Table~\ref{tab:code-size}, other compiler statistics are listed in
Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
Figure~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time}. The benchmark 464.h264ref caused an internal
compiler error with the trunk clang so it could not be reported.

@@

We see some code size reduction in 400.perlbench, 403.gcc, and 453.povray as
shown in Table~\ref{tab:code-size}.  The code size listed here is the text size
delta (from size command in linux) of the final executable for each benchmark.  On
the other hand 447.dealII increased in code size. This is because once the
code-size of a function decreases (due to \GCM{}) it becomes cheaper to do some
optimizations like inlining, loop unrolling, copy propagation, rematerialization
etc. We tried to minimize the noise by disabling inliner, and loop unroller but
still we can see some increase in the code size.  To see the impact of \GCM{} on
inliner we ran \GCM{} with inliner enabled and collected the compile-time
statistics produced by LLVM (See Table~\ref{tab:stats}) . Except for 447.dealII,
450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got more
functions inlined. Also, since the pass runs early, it affects many
optimizations which rely on the number of instructions, length of the use-def
chain, and other metrics.

@@

\section{Conclusions and Future Work}
\label{sec:future-work}
We have presented a fast algorithm to perform \gcm{} of congruent computations
in SSA form. We introduced data structures like $\chi$ nodes, $\Phi$ nodes which
helps exploit flow of values in sparse representation.  Experimental results
show that it is a useful performance and code-size optimization. It also enables
inlining and LICM opportunities, reduces register spill. To preserve performance
and not hoist/sink too much, we have implemented a register pressure aware cost
model as described in Section~\ref{subsec:cost-models}. A part of our
implementation is merged in LLVM trunk as GVNHoist.cpp and the \GCM{}
implementation is under review. This pass does not hoist fully redundant
expressions because they are already handled by a later partial redundancy
elimination pass.

>>>>

  for each $\chi$-nodes in the CFG:
    for each $\chi$-arg in $\chi$-node:
      Remove $\chi$-arg if it fails legality checks

  for each $\chi$-nodes in the CFG:
    if $\chi$ does not have missing entries:
      Perform code hoisting
      update data structures and statistics

@@

  for each VN in VNs
    for each instruction with the same VN
      Find its dominance-frontier (DF)
      Insert the $\Phi$-node at DF

  for each $\Phi$-nodes in the CFG:
    for each $\Phi$-arg in PHI-node:
      Remove $\Phi$-arg if it fails legality checks

  for each $\Phi$-nodes in the CFG:
    if $\Phi$ does not have missing entries:
      Perform code sinking
      if a $\phi$ has same entries, reassign
      update data structures and statistics

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
For evaluation of \gcm{}, we built \SPEC{} with and without the \GCM{}
optimization. All the experiments were conducted on an \xlinux{} machine and at
-Ofast optimization level with inlining and loop unrolling disabled; inliner and
loop-unroller vary wildly with \GCM{} optimization so it was difficult to
visualize the impact. Each benchmark was run three times and the best result was
taken. We collected execution time and code-size which is listed in
Table~\ref{tab:code-size}, other compiler statistics are listed in
Table~\ref{tab:stats}, Table~\ref{tab:spills}, and
Figure~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time}. The benchmark 464.h264ref caused an internal
compiler error with the trunk clang so it could not be reported.

@@

We see some code size reduction in 400.perlbench, 403.gcc, and 453.povray as
shown in Table~\ref{tab:code-size}.  The code size listed here is the text size
delta (from size command in linux) of the final executable for each benchmark.  On
the other hand 447.dealII increased in code size. This is because once the
code-size of a function decreases (due to \GCM{}) it becomes cheaper to do some
optimizations like inlining, loop unrolling, copy propagation, rematerialization
etc. We tried to minimize the noise by disabling inliner, and loop unroller but
still we can see some increase in the code size.

To see the impact of \GCM{} on inliner we ran \GCM{} with inliner enabled and
collected the compile-time statistics produced by LLVM (See
Table~\ref{tab:stats}) . Except for 447.dealII, 450.soplex, 482.sphinx3 and
483.xalancbmk all other benchmarks got more functions inlined. Also, since the
pass runs early, it affects many optimizations which rely on the number of
instructions, length of the use-def chain, and other metrics.

@@

\section{Conclusions and Future Work}
\label{sec:future-work}
We have presented a fast algorithm to perform \gcm{} of congruent computations
in SSA form. We introduced data structures like $\chi$ nodes, $\Phi$ nodes which
helps exploit flow of values in sparse representation.  Experimental results
show that it is a useful performance and code-size optimization. It also enables
inlining and LICM opportunities, reduces register spill. To preserve performance
and not hoist/sink too much, we have implemented a register pressure aware cost
model as described in Section~\ref{subsec:cost-models}. A part of our
implementation is merged in LLVM trunk as GVNHoist.cpp and the \GCM{}
implementation is under review. This pass does not hoist fully redundant
expressions because they are already removed by a later partial redundancy
elimination pass in LLVM.
